---
title: Aesthetic Mapping Exercise
author: Claus Wilke remixed by Joseph Elsherbini
description: Aesthetic Mapping Exercise
date: '2022-08-24'
slug: aesthetic-mapping-exercise
categories: ['exercise']
math: true
output:
  tufte::tufte_html:
    toc: true
    fig_width: 6
    dev: "svg"
---

```{r setup, include=FALSE}
library(purrr)
library(tufte)
library(tidyverse)
library(lubridate)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, eval = FALSE,  comment = "", cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)

# data prep

```

## Introduction

In this worksheet, we will discuss a core concept of ggplot, the mapping of data values onto aesthetics.

We will be using the R package **tidyverse**, which includes `ggplot()` and related functions.

Make a new Rmarkdown document called aesthetic_mapping_excercise.Rmd, and copy the following code chunk:


```{r library-calls, eval=TRUE, echo = FALSE}
# load required library
library(tidyverse)

temperatures <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  mutate(
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(location, day_of_year, month, temperature)

temps_houston <- filter(temperatures, location == "Houston")

```

<pre class="r">
<code class="language-r">
`r ''````{r}
library(tidyverse)

temperatures <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  mutate(
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(location, day_of_year, month, temperature)

temps_houston <- filter(temperatures, location == "Houston")
```
</code>
</pre>

The dataset we will be working with contains the average temperature for each day of the year for Houston, TX:

Whenever you see a code chunk, copy it into your document, and if there are blanks try filling them in to complete the prompt. You can also copy the prose between code chunks, or add your own notes.

```{r temps-houston, eval=TRUE, echo = TRUE}
temps_houston
```

## Basic use of ggplot

In the most basic use of ggplot, we call the `ggplot()` function with a dataset and an aesthetic mapping (created with `aes()`), and then we add a geom, such as `geom_line()` to draw lines or `geom_point()` to draw points.

Try this for yourself. Map the column `day_of_year` onto the x axis and the column `temperature` onto the y axis, and use `geom_line()` to display the data.



```{r ggplot}
ggplot(temps_houston, aes(x = ___, y = ___)) +
  ___()
```


Try again. Now use `geom_point()` instead of `geom_line()`.


```{r ggplot2}
ggplot(temps_houston, aes(x = day_of_year, y = temperature)) +
  ___()
 ```


And now swap which column you map to x and which to y.

  ```{r ggplot3}
  ggplot(temps_houston, aes(x = ___, y = ___)) +
    geom_point()
  ```

## More complex geoms

You can use other geoms to make different types of plots. For example, `geom_boxplot()` will make boxplots. For boxplots, we frequently want categorical data on the x or y axis. For example, we might want a separate boxplot for each month. Try this out. Puth `month` on the x axis, `temperature` on the y axis, and use `geom_boxplot()`.

```{r ggplot-boxplot}
ggplot(temps_houston, aes(x = ___, y = ___)) +
  ___()
```

Now put the month on the y axis and the temperature on the x axis.

```{r ggplot-boxplot2}
ggplot(___) +
  ___()
```

## Adding color

Next we will be working with the dataset `temperatures`, which is similar to `temps_houston` but contains data for three more locations:
    
```{r temperatures, eval=TRUE, echo = TRUE}
temperatures
```

Make a line plot of `temperature` against `day_of_year`, using the `color` aesthetic to color the lines by location. 

```{r ggplot-color}
ggplot(temperatures, aes(x = ___, y = ___, color = ___)) +
  ___()
```


Try again, this time using `location` as the location along the y axis and `temperature` for the color. This plot requires `geom_point()` to look good.

```{r ggplot-color2}
ggplot(___) +
  ___()
```

(Hint: Try `geom_point(size = 5)` to create larger points.)

## Using the `fill` aesthetic

Some geoms use a `fill` aesthetic, which is similar to `color` but applies to shaded areas. (`color` applies to lines and points.) For example, we can use the `fill` aesthetic with `geom_boxplot()` to color the interior of the box. Try this yourself. Plot `month` on x, `temperature` on y, and color the interior of the box by location.

```{r ggplot-fill}
ggplot(temperatures, ___) +
  ___()
```

Can you color the lines of the boxplot by location and the interior by month? Try it.

```{r ggplot-color-fill}
ggplot(temperatures, ___) +
  geom_boxplot()
```

## Using aesthetics as parameters

Many of the aesthetics (such as `color`, `fill`, and also `size` to change line size or point thickness) can be used as parameters inside a geom rather than inside an `aes()` statement. The difference is that when you use an aesthetic as a parameter, you specify a specific value, such as `color = "blue"`, rather than a mapping, such as `aes(color = location)`. Notice the difference: Inside the `aes()` function, we don't actually specify the specific color values, ggplot does that for us. We only say that we want the data values of the `location` column to correspond to different colors. (We will learn later how to tell ggplot to use specific colors in this mapping.)

Try this with the boxplot example from the previous section. Map `location` onto the `fill` aesthetic but set the color of the lines to `"navyblue"`.

```{r ggplot-params}
ggplot(temperatures, ___) +
  ___(___)
```

Now do the reverse. Map `location` onto the line colors but fill the box with the color `"navyblue"`.

```{r ggplot-params2}
ggplot(temperatures, ___) +
  ___(___)
```

