---
title: Figure Design Exercise
author: Claus Wilke remixed by Joseph Elsherbini
description: Figure Design Exercise
date: '2022-08-26'
slug: figure-design-exercise
categories: ['exercise']
math: true
output:
  tufte::tufte_html:
    toc: true
    fig_width: 6
    dev: "svg"
---

```{r setup, include=FALSE}
library(purrr)
library(tufte)
library(tidyverse)
library(palmerpenguins)
library(cowplot)
library(lubridate)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, eval = FALSE,  echo=TRUE, comment = "", cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)


temperatures <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  mutate(
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(location, day_of_year, month, temperature)

temps_months <- read_csv("https://wilkelab.org/SDS375/datasets/tempnormals.csv") %>%
  group_by(location, month_name) %>%
  summarize(mean = mean(temperature)) %>%
  mutate(
    month = factor(
      month_name,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    ),
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(-month_name)
```

## Introduction

In this worksheet, we will discuss how to change and customize themes.

We will be using the R package **tidyverse**, which includes `ggplot()` and related functions. We will also be using the packages **cowplot** for themes and the package **palmerpenguins** for the `penguins` dataset.

<pre class="r">
<code class="language-r">
`r ''````{r library-calls}
# load required library
library(tidyverse)
library(cowplot)
library(palmerpenguins)
```</code></pre>

We will be working with the dataset `penguins` containing data on individual penguins on Antarctica.
<pre class="r">
<code class="language-r">
`r ''````{r penguins}
penguins
```</code></pre>

```{r penguins, echo = FALSE, eval=TRUE}
penguins
```

## Ready-made themes

Let's start with this simple plot with no specific styling.

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point(na.rm = TRUE)  # na.rm = TRUE prevents warning about missing values
```</code></pre>

```{r penguins-simple, eval=TRUE, echo=FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point(na.rm = TRUE)  # na.rm = TRUE prevents warning about missing values
```

The default ggplot theme is `theme_gray()`. Verify that adding this theme to the plot makes no difference in the output. Then change the overall font size by providing the theme function with a numeric font size argument, e.g. `theme_gray(16)`.

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-theme-gray}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point(na.rm = TRUE) +
  ___
```</code></pre>


The **ggplot2** package has many built-in themes, including `theme_minimal()`, `theme_bw()`, `theme_void()`, `theme_dark()`. Try these different themes on the above plot. Also try again changing the font size. You can see all themes provided by **ggplot2** here: https://ggplot2.tidyverse.org/reference/ggtheme.html

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-theme-ggplot2}
 # build all the code for this exercise
```</code></pre>


Many other packages also provide themes. For example, the **cowplot** package provides themes `theme_half_open()`, `theme_minimal_grid()`, `theme_minimal_hgrid()`, and `theme_minimal_vgrid()`. You can see all **cowplot** themes here: https://wilkelab.org/cowplot/articles/themes.html


<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-theme-cowplot}
 # build all the code for this exercise
```</code></pre>


Compare the visual appearance of `theme_minimal()` from **ggplot2** to `theme_minimal_grid()` from **cowplot**. What similarities and differences to you notice? Which do you prefer? (There is no correct answer here, just be aware of the differences and of your preferences.)

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-theme-minimal}
 # build all the code for this exercise
```</code></pre>


## Modifying theme elements

You can modify theme elements by adding a `theme()` call to the plot. Inside the `theme()` call you specify which theme element you want to modify (e.g., `axis.title`, `axis.text.x`, `panel.background`, etc) and what changes you want to make. For example, to make axis titles blue, you would write:

<pre class="r">
<code class="language-r">
`r ''````{r theme-example}
theme(
  axis.title = element_text(color = "blue")
)
```</code></pre>

There are many theme settings, and for each one you need to know what type of an element it is (`element_text()`, `element_line()`, `element_rect()` for text, lines, or rectangles, respectively). A complete description of the available options is available at the **ggplot2** website: https://ggplot2.tidyverse.org/reference/theme.html

Here, we will only try a few simple things. For example, see if you can make the legend title blue and the legend text red.

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-legend}
# make the legend title blue and the legend text red
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point(na.rm = TRUE)
```</code></pre>

```{r penguins-simple-legend. eval=TRUE, echo=FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point(na.rm = TRUE)
```


Now color the area behind the legend in `"aliceblue"`. **Hint:** The theme element you need to change is called `legend.background`. There is also an element `legend.box.background` but it is only visible if `legend.background` is not shown, and in the default **ggplot2** themes that is not the case.

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-legend-background}
 # build all the code for this exercise
```</code></pre>

Another commonly used feature in themes are margins. Many parts of the plot theme can understand customized margins, which control how much spacing there is between different parts of a plot. Margins are typically specified with the function `margin()`, which takes four numbers specifying the margins in points, in the order top, right, bottom, left. So, `margin(10, 5, 5, 10)` would specify a top margin of 10pt, a right margin of 5pt, a bottom margin of 5pt, and a left margin of 10pt.

Try this out by setting the legend margin (element `legend.margin`) such that there is no top and no bottom margin but 10pt left and right margin.

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-margin}
 # build all the code for this exercise
```</code></pre>


There are many other things you can do. Try at least some of the following:

- Change the horizontal or vertical justification of text with `hjust` and `vjust`.
- Change the font family with `family`.<sup>1</sup>
- Change the panel grid. For example, create only horizontal lines, or only vertical lines.
- Change the overall margin of the plot with `plot.margin`.
- Move the position of the legend with `legend.position` and `legend.justification`.
- Turn off some elements by setting them to `element_blank()`.

<sup>1</sup> Getting fonts to work well can be tricky in R. Which specific fonts work depends on the graphics device and the operating system. You can try some of the following fonts and see if they work on app.terra.bio: `"Palatino"`, `"Times"`, `"Helvetica"`, `"Courier"`, `"ITC Bookman"`, `"ITC Avant Garde Gothic"`, `"ITC Zapf Chancery"`.

## Writing your own theme

You can write a theme by 

<pre class="r">
<code class="language-r">
`r ''````{r echo = TRUE, eval=TRUE}
theme_colorful <-
  theme_bw() +
  theme(
    text = element_text(color = "mediumblue"),
    axis.text = element_text(color = "springgreen4"),
    legend.text = element_text(color = "firebrick4")
  )
```</code></pre>

```{r echo = FALSE, eval=TRUE}
theme_colorful <-
  theme_bw() +
  theme(
    text = element_text(color = "mediumblue"),
    axis.text = element_text(color = "springgreen4"),
    legend.text = element_text(color = "firebrick4")
  )
```


**Hint:** Do you have to add `theme_colorful` or `theme_colorful()`? Do you understand which option is correct and why? If you are unsure, try both and see what happens.

<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-theme-colorful}
 # build all the code for this exercise
```</code></pre>


Now write your own theme and then add it to the penguing plot.


<pre class="r">
<code class="language-r">
`r ''````{r penguins-simple-theme-box}
 # build all the code for this exercise
```</code></pre>