---
title: Compound Figures Exercise
author: Claus Wilke remixed by Joseph Elsherbini
description: Compound FiguresExercise
date: '2022-08-26'
slug: compound-figures-exercise
categories: ['exercise']
math: true
output:
  tufte::tufte_html:
    toc: true
    fig_width: 6
    dev: "svg"
---

```{r setup, include=FALSE}
library(purrr)
library(tufte)
library(tidyverse)
library(patchwork)
library(cowplot)
library(lubridate)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, eval = FALSE,  echo=TRUE, comment = "", cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)

p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp))

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear))

p3 <- ggplot(mtcars) + 
  geom_smooth(aes(disp, qsec))

p4 <- ggplot(mtcars) + 
  geom_bar(aes(carb))
```

## Introduction

In this worksheet, we will discuss how to combine several **ggplot2** plots into one compound figure.

We will be using the R package **tidyverse**, which includes `ggplot()` and related functions. We will also be using the package **patchwork** for plot composition.

```{r library-calls, echo = TRUE, eval = FALSE}
# load required library
library(tidyverse)
library(patchwork)
```

We will be working with the dataset `mtcars`, which contains fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

```{r temperatures, eval=TRUE, echo = TRUE}
mtcars
```

## Combining plots

First we set up four different plots that we will subsequently combine. The plots are stored in variables `p1`, `p2`, `p3`, `p4`.

```{r singlet-plots, echo = TRUE, message = FALSE, fig.width = 4, fig.height = 0.618*4}
p1 <- ggplot(mtcars) + 
  geom_point(aes(mpg, disp))
p1  

p2 <- ggplot(mtcars) + 
  geom_boxplot(aes(gear, disp, group = gear))
p2

p3 <- ggplot(mtcars) + 
  geom_smooth(aes(disp, qsec))
p3

p4 <- ggplot(mtcars) + 
  geom_bar(aes(carb))
p4
```

To show plots side-by-side, we use the operator `|`, as in `p1 | p2`. Try this by making a compound plot of plots `p1`, `p2`, `p3` side-by-side.

```{r plots-side-by-side, exercise = TRUE, message = FALSE}
 # build all the code for this exercise
```


To show plots on top of one-another, we use the operator `/`, as in `p1 / p2`. Try this by making a compound plot of plots `p1`, `p2`, `p3` on top of each other.

```{r plots-on-top, exercise = TRUE, message = FALSE}
 # build all the code for this exercise
```


We can also use parentheses to group plots with respect to the operators `|` and `/`. For example, we can place several plots side-by-side and then place this entire row of plots on top of another plot. Try putting `p1`, `p2`, `p3`, on the top row, and `p4` on the bottom row.

```{r plots-complex, exercise = TRUE, message = FALSE}
(___) / ___
```

## Plot annotations

The **patchwork** package provides a powerful annotation system via the `plot_annotation()` function that can be added to a plot assembly. For example, we can add plot tags (the labels in the upper left corner identifying the plots) via the plot annotation `tag_levels`. You can set `tag_levels = "A"` to generate tags A, B, C, etc. Try this out.

```{r plots-w-tags, exercise = TRUE, message = FALSE}
(p1 | p2 | p3 ) / p4
```

Try also tag levels such as `"a"`, `"i"`, or `"1"`.

You can also add elements such as titles, subtitles, and captions, by setting the `title`, `subtitle`, or `caption` argument in `plot_annotation()`. Try this out by adding an overall title to the figure from the previous exercise.

```{r plots-w-title, exercise = TRUE, message = FALSE}
 # build all the code for this exercise
```


Also set a subtitle and a caption.

Finally, you can change the theme of all plots in the plot assembly via the `&` operator, as in `(p1 | p2) & theme_bw()`. Try this out.

```{r themeing, exercise = TRUE, message = FALSE}
 # build all the code for this exercise
```

What happens if you write this expression without parentheses? Do you understand why?