---
title: Visualizing Distributions Exercise
author: Claus Wilke remixed by Joseph Elsherbini
description: Visualizing Distributions Exercise
date: '2022-08-24'
slug: visualizing-distributions-exercise
categories: ['exercise']
math: true
output:
  tufte::tufte_html:
    toc: true
    fig_width: 6
    dev: "svg"
---

```{r setup, include=FALSE}
library(purrr)
library(tufte)
library(tidyverse)
library(ggforce)
library(ggridges)
library(lubridate)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, warning = FALSE, message = FALSE, eval = FALSE,  echo=TRUE, comment = "", cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)

# data prep

```

Load both datasets and the libraries we're using in this exercise

```{r library-calls, echo = TRUE, eval = FALSE}
# load required libraries for this worksheet
library(tidyverse)
library(ggforce)
library(ggridges)
```

```{r eval=TRUE}
# data prep
titanic <- read_csv("https://wilkelab.org/SDS375/datasets/titanic.csv") %>%
  select(age, sex, class, survived) %>%
  arrange(age, sex, class)
  
lincoln_temps <- readRDS(url("https://wilkelab.org/SDS375/datasets/lincoln_temps.rds"))
```

## Introduction

In this worksheet, we will discuss how to display distributions of data values using histograms and density plots.


The first dataset we will be working with contains information about passengers on the Titanic, including their age, sex, the class in which they traveled on the ship, and whether they survived or not:
```{r titanic, eval=TRUE, echo = TRUE}
titanic
```

## Histograms

We start by drawing a histogram of the passenger ages (column `age` in the dataset `titanic`). We can do this in ggplot with the geom `geom_histogram()`. Try this for yourself.

```{r titanic-hist}
ggplot(titanic, aes(___)) +
  ___
```


If you don't specify how many bins you want or how wide you want them to be, `geom_histogram()` will make an automatic choice, but it will also give you a warning that the automatic choice is probably not good. Make a better choice by setting the `binwidth` and `center` parameters. Try the values 5 and 2.5, respectively.

```{r titanic-hist2}
ggplot(titanic, aes(age)) +
  geom_histogram(___)
```

Try a few more different binwidths, e.g. 1 or 10. What are good values for `center` that go with these choices?

## Density plots

Density plots are a good alternative to histograms. We can create them with `geom_density()`. Try this out by drawing a density plot of the passenger ages (column `age` in the dataset `titanic`). Also, by default `geom_density()` does not draw a filled area under the density line. We can change this by setting an explicit fill color, e.g. "cornsilk".

```{r titanic-dens}
ggplot(titanic, aes(___)) +
  ___
```


Just like for histograms, there are options to modify how much detail a density plot shows. A small binwidth in a histogram corresponds to a low bandwidth (`bw`) in a density plot and similarly a large binwidth corresponds to a high bandwidth. In addition, you can change the kernel, e.g. `kernel = "rectangular"` or `kernel = "triangular"`. Try this out by using a bandwidth of 1 and a triangular kernel.

```{r titanic-dens2}
ggplot(titanic, aes(age)) +
  geom_density(fill = "cornsilk", ___)
```

Try a few more different bandwidth and kernel choices, e.g. 0.1 or 10, or rectangular or gaussian kernels. How does the density plot depend on these choices?

## Small multiples (facets)

We can also draw separate histograms for passengers meeting different criteria, for example for passengers traveling in the different classes. Whenever we draw multiple plot panels containing the same type of plot but for different subsets of the data, we speak of "small multiples". In ggplot, we generate small multiples with the function `facet_wrap()`. The function `facet_wrap()` takes as its argument a list of data columns to subdivide the data by. This list is provided via the `vars()` function. For example, `vars(class)` means draw a separate panel for each class, `vars(survived)` means draw a separate panel for each survival status, and `vars(class, survived)` means draw a separate panel for each combination of class and survival status.

As an example, the following code generates small multiple histograms by class:
```{r titanic-hist-class, echo = TRUE}
ggplot(titanic, aes(age)) +
  geom_histogram(binwidth = 5, center = 2.5) +
  facet_wrap(vars(class))
```

Now use the same principle to draw facets of histograms by survival status.

```{r titanic-hist-surv, exercise = TRUE}
ggplot(titanic, aes(age)) +
  geom_histogram(binwidth = 5, center = 2.5) +
  ___
```

Now make a plot that breaks down the data by both survival status and class.

```{r titanic-hist-survclass, exercise = TRUE}
ggplot(titanic, aes(age)) +
  geom_histogram(binwidth = 5, center = 2.5) +
  facet_wrap(vars(survived, ___))
```


Now, do the same but drawing density plots rather than histograms.

```{r titanic-dens-survclass, exercise = TRUE}
ggplot(titanic, aes(age)) +
  ___ +
  facet_wrap(vars(survived, ___))
```


## Multiple distributions at once

Now, let's display many distributions at once, using boxplots, violin plots, strip charts, sina plots, and ridgeline plots.

The next dataset we will be working with contains information about the mean temperature for every day of the year 2016 in Lincoln, NE:
```{r lincoln_temps, eval=TRUE, echo = TRUE}
lincoln_temps
```

## Boxplots and violins

We start by drawing the distributions of mean temperatures for each month of the year (columns `month` and `mean_temp` in the dataset `lincoln_temps`), using boxplots. We can do this in ggplot with the geom `geom_boxplot()`. Try this for yourself.

```{r lincoln-box}
ggplot(lincoln_temps, aes(___)) +
  ___
```

Next, do the same but now using violins (`geom_violin()`) instead of boxplots.

```{r lincoln-violin, exercise = TRUE}
 # build all the code for this exercise
ggplot(___) +

```

Customize the violins by trying some of the following:

- Change the fill or outline color.
- Swap the x and y mappings.
- Change the bandwidth (parameter `bw`) or kernel (parameter `kernel`). These parameters work just like in `geom_density()` as discussed in the single density section above.
- Set `trim = FALSE`. What does this do?


## Strip charts and jittering

Both boxplots and violin plots have the disadvantage that they don't show the individual data points. We can show individual data points by using `geom_point()`. Such a plot is called a *strip chart*.

Make a strip chart for the Lincoln temperature data set. Hint: Use `size = 0.75` to reduce the size of the individual points.

```{r lincoln-strip, exercise = TRUE}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  ___
```

Frequently when we make strip charts we want to apply some jitter to separate points away from each other. We can do so by setting the argument `position = position_jitter()` in `geom_point()`.

When using `position_jitter()` we will normally have to specify how much jittering we want in the horizontal and vertical direction, by setting the `width` and `height` arguments: `position_jitter(width = 0.15, height = 0)`. Both `width` and `height` are specified in units representing the resolution of the data points, and indicate jittering in either direction. So, if data points are 1 unit apart, then `width = 0.15` means the jittering covers 0.3 units or 30% of the spacing of the data points.

Try this for yourself, by making a strip chart with jittering.

```{r lincoln-strip-jitter, exercise = TRUE}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_point(
    size = 0.75,
    ___
  )
```


The function `position_jitter()` applies random jittering to the data points, which means the plot looks different each time you make it. (Verify this.) We can force a specific, fixed arrangement of jittering by setting the `seed` parameter. This parameter takes an arbitrary integer value, e.g. `seed = 1234`. Try this out.

```{r lincoln-strip-jitter2, exercise = TRUE}
 # build all the code for this exercise
ggplot(___) +

```


Finally, try to figure out what the parameter `height` does, by setting it to a value other than 0, or by removing it entirely. 

## Sina plots

We can create a combination of strip charts and violin plots by making sina plots, which jitter points into the shape of a violin. We can do this with `geom_sina()` from the **ggforce** package. Try this out.

```{r lincoln-sina, exercise = TRUE}
 # build all the code for this exercise
ggplot(___) +

```

It often makes sense to draw a sina plot on top of a violin plot. Try this out.

```{r lincoln-sina2, exercise = TRUE}
 # build all the code for this exercise
ggplot(___) +

```

Finally, customize the violins by removing the outline and changing the fill color.

```{r lincoln-sina3, exercise = TRUE}
 # build all the code for this exercise
ggplot(___) +

```


## Ridgeline plots

As the last alternative for visualizing multiple distributions at once, we will make ridgeline plots. These are multiple density plots staggered vertically. In ridgeline plots, we normally map the grouping variable (e.g. here, the month) to the y axis and the dependent variable (e.g. here, the mean temperature) to the x axis.

We can create ridgeline plots using `geom_density_ridges()` from the **ggridges** package. Try this out. Use the column `month_long` instead of `month` for the name of the month to get a slightly nicer plot. Hint: If you get an error about a missing y aesthetic you need to swap your x and y mappings.

```{r lincoln-ridges, exercise = TRUE}
 # build all the code for this exercise
ggplot(___) +

```

What happens when you use `month` instead of `month_long`? Can you explain why?

It is often a good idea to prune the ridgelines once they are close to zero. You can do this with the parameter `rel_min_height`, which takes a numeric value relative to the maximum height of any ridgeline anywhere in the plot. So, `rel_min_height = 0.01` would prune all lines that are less than 1% of the maximum height in the plot.

```{r lincoln-ridges2, exercise = TRUE}
ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges(___)
```

